---
title: "Practical Machine Learning Project"
author: "BJH"
date: "December 23, 2015"
output: 
  html_document: 
    theme: cosmo
---

### Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

#### Data 
The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 

**The goal of this project is to predict the manner in which they did the exercise. i.e. Classe**

### Data Loading and Data Set Preparation

Load the data and necessary packages.

```{r, results = 'hide'}
# Load Known Packages
library(ggplot2)
library(caret)

# Set working directory
setwd("~/Documents/Berlyn_Coursera")

# read Training data file
training <- read.csv("./pml-training.csv", sep=",", na.strings = "NA")
# read Testing data file
testing <- read.csv("./pml-testing.csv", sep=",", na.strings = "NA")
```

Split training set for cross validation.

```{r}
# Set seed
set.seed(654)

# Partition training set
train <- createDataPartition(training$classe, p = 0.7, list = FALSE)
training_split <- training[train, ]
testing_split <- training[-train, ]
```


Next, data is cleaned by removing NAs. Factors such as descriptives (user name, timestamps, etc.) and near zero variables are also removed.


```{r}
# remove near zero values, NAs, and descriptive factors

# remove near zero variance features
nsv <- nearZeroVar(training_split)
training_split <- training_split[, -nsv]

# remove NA vals & descriptive vars

# count missing values
cntlength <- sapply(training_split, function(x) { 
  sum(!(is.na(x) | x == "")) 
  })

# check if missing values total less than 60%
NAcols <- names(cntlength[cntlength < 0.6 * length(training_split$classe)])

# list descriptive columns
descriptivecols <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window")

# columns to include are descriptive and the NAs
excludedcols <- c(descriptivecols, NAcols)

# save training_split with removed values
training_split <- training_split[, !names(training_split) %in% excludedcols]
```

### Prediction Model Generation and Testing
Create model using cleaned data and random forest.

```{r, cache = TRUE}
# load random forest and other required packages
library(randomForest)
library(e1071)

# run random forest
random_forest_model <- randomForest(classe ~ ., data = training_split, importance = TRUE, ntrees = 10)
```

Use the prediction model on the training set and the cross validation set.

```{r}
# run the prediction on the training set used to create the model
ptraining <- predict(random_forest_model, training_split)
print(confusionMatrix(ptraining, training_split$classe))

# run the prediction on the cross validation set
pvalidation <- predict(random_forest_model, testing_split)
print(confusionMatrix(pvalidation, testing_split$classe))

# check the actual distribution of classes amongst the test set
table(testing_split$classe)
```



**The accuracy of the cross validation set is 99.51%.  Therefore, the out of sample error is only 0.49%, indicating a good prediction model.**


### Results for Project Test Set
Write and save answers for official test set.
```{r}
# predict answers
ptest <- predict(random_forest_model, testing)
ptest

# save as...
answers <- as.vector(ptest)

pml_write_files = function(x) {
  n = length(x)
  for (i in 1:n) {
    filename = paste0("problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
                col.names = FALSE)
  }
}

pml_write_files(answers)
```